---
title: "Assessment of Mental Health Need Before, During, and After COVID-19"
author: "Jensen Hu"
always_allow_html: yes
output:
  
---
## Purpose
The purpose of this assessment is to identify areas of potentially increased mental health need as a result of COVID-19 in NYC. The resulting aim is to route public health efforts and mental health resources to those identified areas of highest need.

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE)
# Load packages
library(readr) # input/output
library(dplyr) # data manipulation
library(tidyverse) # data manipulation
library(tigris) # geospatial data
library(sp) # visualization
library(htmlwidgets) # visualization
library(leaflet) # visualization

# Load data 
ds_svi <- read_csv("../RawData/Social Vulnerability.csv", col_types=cols("FIPS"=col_character()))
ds_county_svi <- read_csv("../RawData/svi_ny_county.csv", col_types=cols("FIPS"=col_character()))
nyc_zips <- read_csv("../RawData/nyc_zips.csv")
child_opp <- read_csv("../RawData/ny_child_opp.csv")
covid <- read_csv("https://raw.githubusercontent.com/nychealth/coronavirus-data/master/tests-by-zcta.csv")
covid_boro <- read_csv("https://raw.githubusercontent.com/nychealth/coronavirus-data/master/boro.csv")
```
## Background of Datasets
**For this assessment, we will use the CDC's Social Vulnerability Index along with the child opportunity index to establish the baseline mental health need in NYC. COVID19 related data will then be added onto this baseline as a "COVID19 Layer" to identify those areas of increased mental health need in the context of COVID19**

#### **Social Vulnerability Index (SVI)**   

*  **Background**: In the context of the global COVID-19 pandemic, "SVI data can be used to identify communities that will need continued support to recover following an emergency or natural disaster, and allocate emergency preparedness funding by community need." As seen in current events, COVID19 disproportionately affects Black and Hispanic communities. It is evident that public health institutions not only factor in, but lead with the social determinants of health -- which contribute to racial inequities in health-- when planning and executing emergency resource allocation during and after COVID19.  

#### **Exploration of Social Vulnerability Index (SVI)**

```{r, cache=TRUE, echo=FALSE}
# select relevant svi variables
col_svi <- c("STCNTY","COUNTY","FIPS","RPL_THEME1","RPL_THEME2","RPL_THEME3","RPL_THEME4","RPL_THEMES")
ds_svi <- ds_svi[col_svi]
# bring in NYS census tract spatial data
options(tigris_use_cache = TRUE)
tracts <- tracts(state = 'NY', cb=TRUE)
# merge svi and spatial ny data to create a enriched spatial dataset
svi_tracts <- geo_join(tracts, ds_svi, "GEOID", "FIPS")
# nyc counties only
counties_nyc <- c("New York", "Kings", "Bronx", "Richmond", "Queens")
svi_tracts <- svi_tracts[svi_tracts@data$COUNTY %in% counties_nyc,]
```

SVI consists of 15 census variables organized into 4 different themes: Socioeconomic Status (RPL_THEME1), Household Composition & Disability (RPL_THEME2), Minority Status & Language (RPL_THEME3), Housing & Transportation (RPL_THEME4) and an overall summary (RPL_THEMES). These themes are measured in percentile rank. A higher percentile rank represents greater vulnerability, with a percentile rank of 0.00 meaning the least vulnerable and 1.00 meaning the most vulnerable. Reference:https://svi.cdc.gov/Documents/Data/2016_SVI_Data/SVI2016Documentation.pdf 

These rankings have been alloted by census tract (GEOID)-- "uniquely numbered subdivisions of a county, representing a neighborhood averaging about 4,000 inhabitants." NYC has 5 different counties ("New York", "Kings", "Bronx", "Richmond", "Queens") which have been subdivided into 2,166 census tracts. To provide a frame of reference, census tracts are more granular than zip codes. There are 178 zip codes in NYC.

To begin, we will review SVI data by NY county.

##### **Review of SVI Data by NYC County**

```{r, cache=TRUE, echo=FALSE}
# nyc county fips only
nyc_cfips<-c("36005", "36047","36061", "36085", "36081" )
svi_county<-ds_county_svi%>%filter(FIPS %in% nyc_cfips)
# display table for review
knitr::kable(svi_county[c(4:6, 79,85, 89, 96, 98)], format="markdown")
```

At a glance, the Bronx county has the highest overall social vulnerability (RPL_THEMES=1.000) followed by Kings county (RPL_THEMES=0.918). This seems to indicate that these two counties are likely to be substantially impacted by COVID19 and need priority allocation of emergency resources. 

##### **Mapping SVI data by County**
```{r, echo=FALSE, cache=TRUE}
# bring in NYS county spatial data
county_sp<-counties(state="NY")
# nyc county fips only
county_sp<-county_sp[county_sp@data$GEOID %in% nyc_cfips,]
# svi data merged w/county spatial data
county_merged<-geo_join(county_sp, svi_county, "GEOID", "FIPS")
# pop up
popup <- paste0("County: ", county_merged$COUNTY, "<br>", "SVI: ", county_merged$RPL_THEMES)
# color ranges
pal <- colorNumeric(
  palette = "BuGn",
  domain = county_merged$RPL_THEMES)
# leaflet plot of NYC SVI data by COUNTY
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = county_merged, 
              fillColor = ~pal(county_merged$RPL_THEMES), 
              color = "black", 
              fillOpacity = 0.7, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup) %>%
  addLegend(pal = pal, 
            values = county_merged$RPL_THEMES, 
            position = "bottomright", 
            title = "Social Vulnerability Index")
```

Now, let's take a look at how COVID19 has impacted NYC counties to date:

```{r, echo=FALSE, cache=TRUE}
#
covid_boro<-read_csv("https://raw.githubusercontent.com/nychealth/coronavirus-data/master/boro.csv")
covid_boro<-covid_boro%>%mutate(COUNTY=ifelse(BOROUGH_GROUP=='The Bronx','Bronx', ifelse(BOROUGH_GROUP=='Brooklyn','Kings', ifelse(BOROUGH_GROUP=='Manhattan','New York', ifelse(BOROUGH_GROUP=='Queens', 'Queens', ifelse(BOROUGH_GROUP=='Staten Island','Richmond', 'NA')))))) 
knitr::kable(cc, format="markdown")
```

The above contains cumulative, age adjusted rates of confirmed cases per 100,000 people, by NYC borough of residence. We can see that The Bronx (Bronx County) and Staten Island (Richmond County) have higher COVID case rates in relation to all other counties. Assuming COVID case rate is an indicator of COVID impact, let's add in COVID data into the current SVI rankings. 

Note: I don't think it's this easy.

##### **Adding COVID DATA into Percent Ranking in SVI**
```{r, echo=FALSE}
# keep RPL and SPL columns
svi_county<-svi_county[c(4:6, 79,85, 89, 96:98)]  
# join svi and covid data by county
svi_cc<-left_join(svi_county, covid_boro, by="COUNTY")
# percent rank covid case rate
svi_cc<-svi_cc%>%mutate(covid_rank=percent_rank(COVID_CASE_RATE))
# add percent ranked covid case rate to SPL Themes
svi_cc$adj_SPL<-svi_cc$SPL_THEMES + svi_cc$covid_rank
# percent rank adjusted SPL
svi_cc<-svi_cc%>%mutate(svi_covid_rank=percent_rank(adj_SPL))
colnames(svi_cc)

knitr::kable(svi_cc[c(1,10, 8, 12:15)], format="markdown")
```

Based on the new "svi_covid_rank", we can see that Bronx and Kings county still hold their rank as the highest in social vulnerability in the context of COVID. 

##### **Mapping COVID adjusted SVI data by County**

```{r, echo=FALSE, cache=TRUE}
# join svi-covid data with county spatial data
svi_cc<-geo_join(county_sp, svi_cc, "GEOID", "FIPS")
# pop up
popup <- paste0("County: ", svi_cc$COUNTY, "<br>", "SVI_Covid: ", svi_cc$svi_covid_rank)
# color ranges
pal <- colorNumeric(
  palette = "BuGn",
  domain = svi_cc$svi_covid_rank)
# leaflet plot of NYC SVI data by COUNTY
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = county_merged, 
              fillColor = ~pal(svi_cc$svi_covid_rank), 
              color = "black", 
              fillOpacity = 0.7, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup) %>%
  addLegend(pal = pal, 
            values = svi_cc$svi_covid_rank, 
            position = "bottomright", 
            title = "COVID adjusted Social Vulnerability Index")

```

#### ZIP CODE CLEAN/TRANSFORM #178 zip codes in nyc

```{r, echo=FALSE}
# 178 zip codes in nyc - separate comma sep values into individual rows
nyc_zips<-separate_rows(nyc_zips, `ZIP Codes`, convert=FALSE)
colnames(nyc_zips)[3]<-"ZIP"
# use ZCTAS and nyc zips to create NYC ZIP spatial layer
options(tigris_use_cache = TRUE)
zctas_ny<-zctas(cb=TRUE, starts_with=c("10","11"))
x<-nyc_zips$ZIP
nyc_zips<-zctas_ny[zctas_ny@data$ZCTA5CE10 %in% x, ]
# NYC ZIP SPATIAL+ COVID-19 Data
nyc_covid<-geo_join(nyc_zips, covid, "ZCTA5CE10", "MODZCTA")
```

```{r, echo=FALSE, eval=FALSE}
## select child opp variables
y<-c(2,3,6,7,15:18)
child_opp<-child_opp[y]
colnames(child_opp)[1]<-"GEOID"
# set factor order in plot value
child_opp$c5_COI_nat<-as.factor(child_opp$c5_COI_nat)
child_opp$c5_COI_nat<-ordered(child_opp$c5_COI_nat, levels=c("Very Low", "Low", "Moderate", "High", "Very High"))
# enrich geo spatial data w/child opp data
merge_opp<-geo_join(tracts, child_opp, "GEOID", "GEOID")
# only NYC County FIPS
merge_opp<-merge_opp[merge_opp@data$countyfips %in% nyc_cfips,]
```


#### Leaflet#1 - SVI by Census Tract

```{r , echo=FALSE}
# popup
popup1 <- paste0("GEOID: ", svi_tracts$GEOID, "<br>", "Social Vulnerability Index: ", svi_tracts$RPL_THEMES,"<br>", "COUNTY: ", svi_tracts$COUNTY)
# pal
pal1 <- colorNumeric(
  palette = "YlGnBu",
  domain = svi_tracts$RPL_THEMES)
## leaflet plot of NYC SVI data by TRACT
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = svi_tracts, 
              fillColor = ~pal1(RPL_THEMES), 
              color = "#b2aeae", # you need to use hex colors
              fillOpacity = 0.7, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup1) %>%
  addLegend(pal = pal1, 
            values = svi_tracts$RPL_THEMES, 
            position = "bottomright", 
            title = "Social Vulnerability Index")
            #labFormat = labelFormat(suffix = "%")) 
```

#### Leaflet#2 - NYC ZIPS
```{r, echo=FALSE}
# leaflet plot of NYC ZIP LAYER
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = nyc_zips,
              weight =1,
              color= "black",
              fillOpacity = 0.1)
```
#### Leaflet#3 - NYC ZIPS + COVID19 % Positive
```{r, echo=FALSE}
# hover tools 
popup3 <- paste0("Zip Code: ", nyc_covid$ZCTA5CE10, "<br>", "Percent Positive Cases: ", nyc_covid$zcta_cum.perc_pos)
# color ranges
pal3 <- colorNumeric(
  palette = "YlGnBu",
  domain= nyc_covid$zcta_cum.perc_pos)
# NYC ZIPS + COVID19 % Positive
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = nyc_covid,
              fillColor = ~pal3(zcta_cum.perc_pos),
              weight =2,
              color = "black",
              fillOpacity = 0.7,
              popup = popup3)%>%
    addLegend(pal = pal3, 
            values = nyc_covid$zcta_cum.perc_pos, 
            position = "bottomright", 
            title = "Percent Positive",
            labFormat = labelFormat(suffix = "%")) 
```
#### Leaflet#4 - NYC CHILD OPP
```{r , echo=FALSE}
##LEAFLET PLOTTING
#hover tools 
popup2 <- paste0("GEOID: ", merge_opp$GEOID, "<br>", "Child Opportunity Level: ", merge_opp$c5_COI_nat,"<br>", "COUNTY: ", merge_opp$msaname15)
#color ranges
pal2 <- colorFactor(
  palette = "YlGnBu",
  domain = merge_opp$c5_COI_nat,
  reverse = TRUE)
#leaflet plot of NYC CHILD OPP data by TRACT
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = merge_opp, 
              fillColor = ~pal2(c5_COI_nat), 
              color = "#b2aeae", # you need to use hex colors
              fillOpacity = 0.7, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup2) %>%
  addLegend(pal = pal2, 
            values = merge_opp$c5_COI_nat, 
            position = "bottomright", 
            title = "Child Opp")
#labFormat = labelFormat(suffix = "%")) 
```
#### Leaflet#5 - SVI by Census Tract + ZCTAS (ZIP) Overlay
```{r , echo=FALSE}
##LEAFLET PLOTTING
#hover tools 
popup1 <- paste0("GEOID: ", svi_tracts$GEOID, "<br>", "Social Vulnerability Index: ", svi_tracts$RPL_THEMES,"<br>", "COUNTY: ", svi_tracts$COUNTY)

#color ranges
pal1 <- colorNumeric(
  palette = "YlGnBu",
  domain = svi_tracts$RPL_THEMES)

#Attempted combination
#leaflet plot of NYC CHILD OPP data by TRACT
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
#  addPolygons(data = merge_opp, 
#              fillColor = ~pal2(c5_COI_nat), 
#              color = "#b2aeae", # you need to use hex colors
#              fillOpacity = 0.9, 
#              weight = 1, 
#              smoothFactor = 0.2,
#              popup = popup2) %>%
  addPolygons(data = svi_tracts, 
              fillColor = ~pal1(RPL_THEMES), 
              color = "#b2aeae", # you need to use hex colors
              fillOpacity = 0.9, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup1) %>%
  addPolygons(data = nyc_zips,
              weight =1,
              color= "black",
              fillOpacity = 0.1)%>%
#  addLegend(pal = pal2, 
#            values = merge_opp$c5_COI_nat, 
#            position = "bottomright", 
#            title = "Child Opp")%>%
  addLegend(pal = pal1, 
            values = svi_tracts$RPL_THEMES, 
            position = "bottomleft", 
            title = "SVI")
```
